Proofs for ``Geometry of Local Spectral Clustering''

### 2021-01-27 jmlr/revision/proofs.tex
Editing pass over the main document complete.

The proofs of our major theorems largely consist of (at most) three modular parts.
\begin{enumerate}
	\item \textbf{Fixed graph results.} Results which hold with respect to an arbitrary graph $G$, and are stated with respect to functionals (i.e. normalized cut, conductance, and local spread) of $G$;
	\item \textbf{Sample-to-population results.} For the specific choice of $G = G_{n,r}$, we relate the aforementioned functionals to their population analogues. This proves~\textcolor{red}{(?)} 
	\item \textbf{Bounds on population functionals.} (In the case of density clustering only.) When the candidate cluster is a $\lambda$-density cluster, we bound the population functionals by a function of $\lambda$, as well as the other relevant parameters introduced in Section~\textcolor{red}{(?)}. This proves~\textcolor{red}{(?)}
\end{enumerate}
The first three sections of this appendix will correspond to each of these three parts. In Section~\ref{sec:pf_major_theorems}, we will combine these parts to prove the major theorems of our main text, \textcolor{red}{(...)}. Finally, in Section~\ref{sec:appr_misclassification_error} we derive upper bounds for the aPPR vector, and in Section~\ref{sec:experimental_details} we give relevant details regarding our experiments.

@@

\paragraph{Organization.} In the following subsections we establish: (\ref{subsec:pf_lem_zhu}) an upper bound on the misclassification error of PPR in terms of $\alpha$ and $\Phi(C;G)$ (Lemma~\ref{lem:zhu}), and an analogous result for aPPR (Corollary~\ref{cor:zhu}); (\ref{subsec:ppr_uniform_bounds}) a uniform bound on the perturbations of the PPR vector, to be used later in the proof of Theorem~\ref{thm:density_cluster_consistent_recovery} (consistency of PPR);  (\ref{subsec:lovasz_simonovits_bounds}) upper bounds on the mixedness of $q_v^{(t)}$ (as a function of $t$) and $p_v$ (as a function of $\alpha$), which will be helpful in the proofs of Proposition~\ref{prop:pointwise_mixing_time} and Theorem~\ref{thm:ppr_lb}; (\ref{subsec:pf_prop_pointwise_mixing_time}) an upper bound on $\tau_{\infty}(G)$ in terms of $\Psi(G)$ and $s(G)$ (Proposition~\ref{prop:pointwise_mixing_time}); and (\ref{subsec:ppr_spectral_partitioning}) an upper bound on the normalized cut $\Phi(\wh{C};G)$ in terms of $\Phi(C;G)$, to be used later in the proof of Theorem~\ref{thm:ppr_lb} (negative example).

@@

The proof of Proposition~\ref{prop:pointwise_mixing_time} is then straightforward.
\begin{proof}[Proof (of Proposition~\ref{prop:pointwise_mixing_time})]
	Put $t_{\ast} = 8/(\Psi(G)^2) \ln(4/s(G)) + 4$. We will use Theorem~\ref{thm:mixing_time_rw} to show that $\mathrm{TV}(q_v^{(t_{\ast})},\pi) \leq 1/4$. This will in turn imply (\cite{montenegro2002} pg. 13) that for $\tau_{\ast} = t_{\ast} \log_2(8/s(G))$,
	\begin{equation*}
	\mathrm{TV}(q_v^{(\tau_{\ast})},\pi) \leq \frac{1}{8}s(G),
	\end{equation*}
	and applying Lemma~\ref{lem:tv_to_pointwise} gives
	\begin{equation*}
	\max_{u \in V} \Bigl\{\frac{\pi(u) - q_v^{(\tau_{\ast} + 1)}(u)}{\pi(u)}\Bigr\} \leq \frac{1}{4}.
	\end{equation*}
	Taking maximum over all $v \in V$, we conclude that $\tau_{\infty}(G) \leq \tau_{\ast} + 1$, which is exactly the claim of Proposition~\ref{prop:pointwise_mixing_time}.

@@

\subsection{Spectral partitioning properties of PPR}
\label{subsec:ppr_spectral_partitioning}
The following theorem is the main result of~\ref{subsec:ppr_spectral_partitioning}. It relates the normalized cut of the sweep sets $\Phi(S_{\beta};G)$ to the normalized cut of a candidate cluster $C \subseteq V$, when $p_v$ is properly initialized within $C$.

@@

\subsection{Review: concentration inequalities}
\label{subsec:concentration}
We use Hoeffding's inequality to control the deviations of the empirical measure of $\mc{S}$.
\begin{lemma}[Hoeffding's Inequality.]
	\label{lem:hoeffding_2}
	Fix $\delta \in (0,1)$. For any measurable $\mc{S} \subseteq \Rd$,
	\begin{equation*}
	(1 - \delta) \Pbb(\mathcal{S}) \leq \Pbb_n(\mathcal{S}) \leq (1 + \delta)\Pbb(\mathcal{S})
	\end{equation*}
	with probability at least $1 - 2\exp\set{-2\delta^2(\Pbb(\mathcal{S}))^2n}$. 
\end{lemma}

Many graph functionals are order-2 U-statistics, and we use Hoeffding's inequality to control the deviations of these functionals from their expectations. Recall that $U_n$ is an order-2 U-statistic with kernel $\varphi: \Rd \times \Rd \to \Reals$ if 
\begin{equation*}
U_n = \frac{1}{n(n-1)}\sum_{i = 1}^{n} \sum_{j \neq i} \varphi(x_i,x_j).
\end{equation*}
We write $\norm{\varphi}_{\infty} = \sup_{x,y} |\varphi(x,y)|$. 
\begin{lemma}[Hoeffding's Inequality for Order-2 U-statistics.]
	\label{lem:hoeffding}
	Fix $\delta \in (0,1)$. Assume $\norm{\varphi}_{\infty} \leq 1$. Then,
	\begin{equation*}
	(1 - \delta) \mathbb{E}[U_n] \leq U_n \leq (1 + \delta) \mathbb{E}[U_n]
	\end{equation*}
	with probability at least $1 - 2 \exp\left(-\delta^2 (\mathbb{E}[U_n])^2 n\right)$.
\end{lemma}

Finally, we use Lemma~\ref{lem:bernstein_union}---a combination of Bernstein's inequality and a union bound---to upper and lower bound $d_{\max}(G_{n,r})$ and $d_{\min}(G_{n,r})$. For measurable sets $\mc{S}_1,\ldots,\mc{S}_M$, we denote $p_{\min} := \min_{m = 1,\ldots,M} \Pbb(\mathcal{A}_m)$, and likewise let $p_{\max} := \max_{m = 1,\ldots,M} \Pbb(\mathcal{A}_m)$
\begin{lemma}[Bernstein's inequality + union bound.]
	\label{lem:bernstein_union}
	Fix $\delta \in (0,1)$. For any measurable $\mathcal{S}_1,\ldots,\mathcal{S}_M \subseteq \Rd$, 
	\begin{equation*}
	(1 - \delta) p_{\min} \leq \min_{m = 1,\ldots,M} \Pbb_n(\mathcal{A}_m) \leq \max_{m = 1,\ldots,M}  \Pbb_n(\mathcal{A}_m) \leq (1 + \delta) p_{\max}
	\end{equation*}
	with probability at least $1 - 2 M \exp\left\{-\frac{\frac{1}{3}\delta^2p_{\min}n}{1 + \frac{\delta}{3}}\right\}$.  
\end{lemma}

\subsection{Sample-to-population: normalized cut}
\label{subsec:sample_to_population_ncut}
In this subsection we establish~\eqref{eqn:sample_to_population_normalized_cut}. For a set $\mc{S} \subseteq \Rd$, both $\mathrm{cut}_{n,r}(\mc{S}[X])$ and $\mathrm{vol}_{n,r}(\mc{S}[X])$ are order-$2$ U-statistics:
\begin{align*}
\mathrm{cut}_{n,r}(\mc{S}[X]) & = \sum_{i = 1}^{n} \sum_{j \neq i} \1\{\|x_i - x_j\| \leq r\} \cdot \1\{x_i \in \mc{S}\} \cdot \1\{x_j \not\in \mc{S}\},
\intertext{and}
\mathrm{vol}_{n,r}(\mc{S}[X]) & = \sum_{i = 1}^{n} \sum_{j \neq i} \1\{\|x_i - x_j\| \leq r\} \cdot \1\{x_i \in \mc{S}\}.
\end{align*}
Therefore with probability at least $1 - \exp\{-\delta^2 (\cut_{\Pbb,r}(\mc{S}))^2\}$,
\begin{equation*}
\frac{1}{n(n - 1)} \mathrm{cut}_{n,r}(\mc{S}[X]) \leq (1 + \delta) \cut_{\Pbb,r}(\mc{S}),
\end{equation*}
and likewise with probability at least $1 - \exp\{-\delta^2 (\vol_{\Pbb,r}(\mc{S}))^2\}$,
\begin{equation*}
(1 - \delta) \vol_{\Pbb,r}(\mc{S}) \leq \frac{1}{n(n - 1)} \vol_{n,r}(\mc{S}[X]).
\end{equation*}
Consequently, for any $\delta \in (0,1/3)$,  
\begin{equation*}
\Phi_{n,r}(\mc{C}[X]) \leq \frac{1 + \delta}{1 - \delta} \cdot  \frac{\cut_{\Pbb,r}(\mc{C})}{\min\{\vol_{\Pbb,r}(\mc{C}), \vol_{\Pbb,r}(\mc{C}^c)\}} = \frac{1 + \delta}{1 - \delta} \cdot \Phi_{\Pbb,r}(\mc{C}) \leq (1 + 3\delta) \cdot \Phi_{\Pbb,r}(\mc{C})  
\end{equation*}
with probability at least $1 - 3\exp\{-\delta^2 (\cut_{\Pbb,r}(\mc{C}))^2\}$. This establishes~\eqref{eqn:sample_to_population_normalized_cut} upon taking $b_1 := (\cut_{\Pbb,r}(\mc{C}))^2$.

\subsection{Sample-to-population: local spread}
\label{subsec:sample_to_population_local_spread}
In this subsection we establish~\eqref{eqn:sample_to_population_local_spread}. To ease the notational burden, let $\wt{G}_{n,r} := G_{n,r}\bigl[\mc{C}[X]\bigr]$. By Lemma~\ref{lem:hoeffding_2}, with probability at least $1 - \exp\{-2\delta^2 (\Pbb(\mc{C}))^2n\}$
\begin{equation}
\label{pf:sample_to_population_local_spread_1}
\wt{n} \geq (1 - \delta) \cdot n \cdot \Pbb(\mc{C}).
\end{equation}
Conditional on~\eqref{pf:sample_to_population_local_spread_1}, it follows from Lemma~\ref{lem:bernstein_union} that with probability at least $1 -n\exp\{-\delta^2d_{\min}(\wt{\Pbb}) \cdot (1 - \delta) \Pbb(\mc{C}) n/(3 + \delta)\}$,
\begin{equation}
\label{pf:sample_to_population_local_spread_2}
\frac{1}{\wt{n}} d_{\min}(\wt{G}_{n,r}) \geq (1 - \delta) \cdot d_{\min}(\wt{\Pbb}),
\end{equation}
and likewise it follows from Lemma~\ref{lem:hoeffding} that with probability at least $1 - \exp\{-\delta^2(\vol_{\wt{\Pbb},r}(\mc{C}))^2 \cdot (1 - \delta) \Pbb(\mc{C}) n \}$,
\begin{equation*}
\frac{1}{\wt{n}(\wt{n} - 1)} \vol(\wt{G}_{n,r}) \geq (1 - \delta) \cdot \vol_{\wt{\Pbb},r}(\mc{C}).
\end{equation*}
Consequently for any $\delta \in (0,1/3)$,
\begin{equation*}
s_{n,r}(\mc{C}[X]) = \frac{d_{\min}(\wt{G}_{n,r})^2}{\vol(\wt{G}_{n,r})} \geq \frac{\frac{1}{\wt{n}^2}d_{\min}(\wt{G}_{n,r})^2}{\frac{1}{\wt{n}(\wt{n} - 1)}\vol(\wt{G}_{n,r})} \geq \frac{(1 - \delta)^2}{(1 + \delta)} \cdot  \frac{d_{\min}(\wt{\Pbb})^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \geq (1 - 3\delta) \cdot s_{\Pbb,r}(\mc{C}).
\end{equation*}
with probability at least $1 - \exp\{-2\delta^2 (\Pbb(\mc{C}))^2n\} - \exp\bigl\{-2\delta^2\bigl(\vol_{\wt{\Pbb},r}(\mc{C})\bigr)^2 \Pbb(\mc{C})n/3\bigr\} - n\exp\bigl\{-\delta^2 d_{\min}(\wt{\Pbb}) \cdot \Pbb(\mc{C}) n/5\bigr\}$. This establishes~\eqref{eqn:sample_to_population_local_spread} upon taking $b_2 := (\Pbb(\mc{C}))^2 \wedge 2\bigl(\vol_{\wt{\Pbb},r}(\mc{C})\bigr)^2 \Pbb(\mc{C})/3 \wedge d_{\min}(\wt{\Pbb}) \cdot \Pbb(\mc{C})/5$.

@@

In this subsection we establish~\eqref{eqn:sample_to_population_conductance}. As mentioned in our main text, the proof of~\eqref{eqn:sample_to_population_conductance} relies on a high-probability upper bound of the $\infty$-transportation distance between $\mbb{P}$ and $\mbb{P}_n$, from~\citep{garciatrillos16b}. We begin by reviewing this upper bound, in Theorem~\ref{thm:garciatrillos16}. Subsequently in Proposition~\ref{prop:conductance_lb_transportation_distance}, we relate the $\infty$-transportation distance between two measures $\mbb{Q}_1$ and $\mbb{Q}_2$ to the difference of their conductances. Together these results will imply~\eqref{eqn:sample_to_population_conductance}.

@@

\paragraph{$\infty$-transportation distance between empirical and population measures.}
Under mild regularity conditions, \cite{garciatrillos16b} upper bound the transportation distance $\Delta_{\infty}(\mbb{P},\mbb{P}_n)$. 
\begin{enumerate}[label=(A\arabic*)]
	\item 
	\label{asmp:domain} 
	The distribution $\Pbb$ is defined on a domain $D \subseteq \Rd$, which is a bounded, connected, open set with Lipschitz boundary. 
	\item 
	\label{asmp:bounded_density} 
	The distribution $\Pbb$ has density $g: D \to (0,\infty)$ such that there exist $g_{\min} \leq 1 \leq g_{\max}$ for which
	\begin{equation*}
	(\forall x \in D)~~ g_{\min} \leq g(x) \leq g_{\max}
	\end{equation*}
\end{enumerate}
\begin{theorem}[Theorem~1.1 of \cite{garciatrillos16}]
	\label{thm:garciatrillos16}
	Suppose $\Pbb$ satisfies~\ref{asmp:domain} and~\ref{asmp:bounded_density}. Then, there exists positive constants $B_{2}$ and $B_4$ that do not depend on $n$, such that with probability at least $1 - B_2/n$:
	\begin{equation*}
	\Delta_{\infty}(\mbb{P},\mbb{P}_n) \leq B_4 \cdot 
	\begin{dcases*}
	\frac{\ln(n)^{3/4}}{n^{1/2}},&~~\textrm{if $d = 2$,} \\
	\frac{\ln(n)^{1/d}}{n^{1/d}},&~~\textrm{if $d \geq 3$.}
	\end{dcases*}
	\end{equation*}
\end{theorem}
Assuming the candidate cluster $\mc{C}$ and conditional distribution $\wt{\Pbb}$ satisfy~\ref{asmp:domain} and~\ref{asmp:bounded_density}, then Theorem~\ref{thm:garciatrillos16} applies to $\Delta_{\infty}(\wt{\Pbb},\wt{\Pbb}_n)$; we will use this upper bound on $\Delta_{\infty}(\wt{\Pbb},\wt{\Pbb}_n)$ to show~\eqref{eqn:sample_to_population_conductance}.

@@

\subsection{Population functionals, hard case}
\label{subsec:density_cluster_hard_case}
Let $\Pbb$ be the hard case distribution over rectangular domain $\mc{X}$, defined as in~\eqref{eqn:lb_density}, and $\mc{L}$ the lower half of $\mc{X}$. Suppose $r \in (0,\sigma/2)$. Then the population normalized cut $\Phi_{\Pbb,r}(\mathcal{L})$ is lower bounded,
\begin{equation}
\label{eqn:ncut_lower_set}
\Phi_{\Pbb,r}(\mc{L}) \leq \frac{8}{3} \cdot \frac{r}{\rho}.
\end{equation}
and the population local spread $s_{\Pbb,r}(\mc{X})$ is upper bounded,
\begin{equation}
\label{eqn:local_spread_ub}
s_{\Pbb,r}(\mc{X}) \geq \frac{\pi r^2 \epsilon^2}{2 \rho\sigma}
\end{equation}

>>>>

The proofs of our major theorems largely consist of (at most) three modular parts.
\begin{enumerate}
	\item \textbf{Fixed graph results.} Results which hold with respect to an arbitrary graph $G$, and are stated with respect to functionals (i.e. normalized cut, conductance, and local spread) of $G$;
	\item \textbf{Sample-to-population results.} For the specific choice $G = G_{n,r}$, we relate the aforementioned functionals to their population analogues. 
	\item \textbf{Bounds on population functionals.} (In the case of density clustering only.) When the candidate cluster is a $\lambda$-density cluster, we bound the population functionals by a function of $\lambda$, as well as the other relevant parameters introduced in Section~\ref{sec:ppr_density_cluster}.
\end{enumerate}
The first three sections of this appendix will correspond to each of these three parts. In Section~\ref{sec:pf_major_theorems}, we will combine these parts to prove the major theorems of our main text, Theorems~\ref{thm:volume_ssd_ub} and~\ref{thm:density_cluster_volume_ssd_ub}, as well as our negative result, Theorem~\ref{thm:ppr_lb}. In Section~\ref{sec:appr_misclassification_error} we derive upper bounds for the aPPR vector, and show that under certain conditions the PPR vector can perfectly separate two density clusters. Finally, in Section~\ref{sec:experimental_details} we give relevant details regarding our experiments.

@@

\paragraph{Organization.} In the following sections we establish: (Section \ref{subsec:pf_lem_zhu}) an upper bound on the misclassification error of PPR in terms of $\alpha$ and $\Phi(C;G)$ (Lemma~\ref{lem:zhu}), and an analogous result for aPPR (Corollary~\ref{cor:zhu}); (Section \ref{subsec:ppr_uniform_bounds}) a uniform bound on the perturbations of the PPR vector, to be used later in the proof of Theorem~\ref{thm:density_cluster_consistent_recovery} (consistency of PPR);  (Section \ref{subsec:lovasz_simonovits_bounds}) upper bounds on the mixedness of $q_v^{(t)}$ (as a function of $t$) and $p_v$ (as a function of $\alpha$), which will be helpful in the proofs of Proposition~\ref{prop:pointwise_mixing_time} and Theorem~\ref{thm:ppr_lb}; (Section \ref{subsec:pf_prop_pointwise_mixing_time}) an upper bound on $\tau_{\infty}(G)$ in terms of $\Psi(G)$ and $s(G)$ (Proposition~\ref{prop:pointwise_mixing_time}); and (Section \ref{subsec:ppr_spectral_partitioning}) an upper bound on the normalized cut $\Phi(\wh{C};G)$ in terms of $\Phi(C;G)$, to be used later in the proof of Theorem~\ref{thm:ppr_lb} (negative example).

@@

The proof of Proposition~\ref{prop:pointwise_mixing_time} is then straightforward.
\begin{proof}[Proof (of Proposition~\ref{prop:pointwise_mixing_time})]
	Put $t_{\ast} = 8/(\Psi(G)^2) \ln(4/s(G)) + 4$. We will use Theorem~\ref{thm:mixing_time_rw} to show that $\mathrm{TV}(q_v^{(t_{\ast})},\pi) \leq 1/4$. This will in turn imply (\cite{montenegro2002} pg. 13) that for $\tau_{\ast} = t_{\ast} \log_2(8/s(G))$,
	\begin{equation*}
	\mathrm{TV}(q_v^{(\tau_{\ast})},\pi) \leq \frac{1}{8}s(G),
	\end{equation*}
	and applying Lemma~\ref{lem:tv_to_pointwise} gives
	\begin{equation*}
	\max_{u \in V} \Bigl\{\frac{\pi(u) - q_v^{(\tau_{\ast} + 1)}(u)}{\pi(u)}\Bigr\} \leq \frac{1}{4}.
	\end{equation*}
	Taking maximum over all $v \in V$, we conclude that $\tau_{\infty}(G) \leq \tau_{\ast} + 1$, which implies the claim of Proposition~\ref{prop:pointwise_mixing_time}.

@@

\subsection{Spectral partitioning properties of PPR}
\label{subsec:ppr_spectral_partitioning}
The following theorem is the main result of Section~\ref{subsec:ppr_spectral_partitioning}. It relates the normalized cut of the sweep sets $\Phi(S_{\beta};G)$ to the normalized cut of a candidate cluster $C \subseteq V$, when $p_v$ is properly initialized within $C$.

@@

\subsection{Review: concentration inequalities}
\label{subsec:concentration}
We use Bernstein's inequality to control the deviations of the empirical probability of $\mc{S}$.
\begin{lemma}[Bernstein's Inequality.]
	\label{lem:hoeffding_2}
	Fix $\delta \in (0,1)$. For any measurable $\mc{S} \subseteq \Rd$, each of the inequalities,
	\begin{equation*}
	(1 - \delta) \Pbb(\mathcal{S}) \leq \Pbb_n(\mathcal{S})~~\textrm{and}~~\Pbb_n(\mathcal{S}) \leq (1 + \delta)\Pbb(\mathcal{S}),
	\end{equation*}
	hold with probability at least $1 - \exp\set{-n\delta^2\Pbb(\mathcal{S})/(2 + 2\delta)} \leq 1 - \exp\set{-n\delta^2\Pbb(\mathcal{S})/4}$. 
\end{lemma}

Many graph functionals are order-2 U-statistics, and we use Bernstein's inequality to control the deviations of these functionals from their expectations. Recall that $U_n$ is an order-2 U-statistic with kernel $\varphi: \Rd \times \Rd \to \Reals$ if 
\begin{equation*}
U_n = \frac{1}{n(n-1)}\sum_{i = 1}^{n} \sum_{j \neq i} \varphi(x_i,x_j).
\end{equation*}
We write $\norm{\varphi}_{\infty} = \sup_{x,y} |\varphi(x,y)|$. 
\begin{lemma}[Bernstein's Inequality for Order-2 U-statistics.]
	\label{lem:hoeffding}
	Fix $\delta \in (0,1)$. Assume $\norm{\varphi}_{\infty} \leq 1$. Then each of the inequalities,
	\begin{equation*}
	(1 - \delta) \mathbb{E}U_n \leq U_n~~\textrm{and}~~ U_n\leq (1 + \delta) \mathbb{E}U_n,
	\end{equation*}
	hold with probability at least $1 - \exp\{-n\delta^2 \Ebb U_n/(4 + 4\delta/3)\} \leq 1 - \exp\{-n\delta^2 \Ebb U_n/6\}$.
\end{lemma}

Finally, we use Lemma~\ref{lem:bernstein_union}---a combination of Bernstein's inequality and a union bound---to upper and lower bound $d_{\max}(G_{n,r})$ and $d_{\min}(G_{n,r})$. For measurable sets $\mc{S}_1,\ldots,\mc{S}_M$, we denote $p_{\min} := \min_{m = 1,\ldots,M} \Pbb(\mathcal{A}_m)$, and likewise let $p_{\max} := \max_{m = 1,\ldots,M} \Pbb(\mathcal{A}_m)$
\begin{lemma}[Bernstein's inequality + union bound.]
	\label{lem:bernstein_union}
	Fix $\delta \in (0,1)$. For any measurable $\mathcal{S}_1,\ldots,\mathcal{S}_M \subseteq \Rd$, each of the inequalities
	\begin{equation*}
	(1 - \delta) p_{\min} \leq \min_{m = 1,\ldots,M} \Pbb_n(\mathcal{A}_m),~~\textrm{and}~~\max_{m = 1,\ldots,M}  \Pbb_n(\mathcal{A}_m) \leq (1 + \delta) p_{\max}
	\end{equation*}
	hold with probability at least $1 - M \exp \{-n\delta^2p_{\min}/(2 + 2\delta) \} \geq 1 - M \exp\{-n\delta^2p_{\min}/4\}$.  
\end{lemma}

\subsection{Sample-to-population: normalized cut}
\label{subsec:sample_to_population_ncut}
In this subsection we establish~\eqref{eqn:sample_to_population_normalized_cut}. For a set $\mc{S} \subseteq \Rd$, both $\mathrm{cut}_{n,r}(\mc{S}[X])$ and $\mathrm{vol}_{n,r}(\mc{S}[X])$ are order-$2$ U-statistics:
\begin{align*}
\mathrm{cut}_{n,r}(\mc{S}[X]) & = \sum_{i = 1}^{n} \sum_{j \neq i} \1\{\|x_i - x_j\| \leq r\} \cdot \1\{x_i \in \mc{S}\} \cdot \1\{x_j \not\in \mc{S}\},
\intertext{and}
\mathrm{vol}_{n,r}(\mc{S}[X]) & = \sum_{i = 1}^{n} \sum_{j \neq i} \1\{\|x_i - x_j\| \leq r\} \cdot \1\{x_i \in \mc{S}\}.
\end{align*}
Therefore with probability at least $1 - \exp\{-n\delta^2\cut_{\Pbb,r}(\mc{S})/4\}$,
\begin{equation*}
\frac{1}{n(n - 1)} \mathrm{cut}_{n,r}(\mc{S}[X]) \leq (1 + \delta) \cut_{\Pbb,r}(\mc{S}),
\end{equation*}
and likewise with probability at least $1 - \exp\{-n\delta^2 \vol_{\Pbb,r}(\mc{S})/4\} - \exp\{-n\delta^2 \vol_{\Pbb,r}(\mc{S}^c)/4\}$,
\begin{equation*}
(1 - \delta) \vol_{\Pbb,r}(\mc{S}) \leq \frac{1}{n(n - 1)} \vol_{n,r}(\mc{S}[X]),~~\textrm{and}~~(1 - \delta) \vol_{\Pbb,r}(\mc{S}^c) \leq \frac{1}{n(n - 1)} \vol_{n,r}(\mc{S}^c[X])
\end{equation*}
Consequently, for any $\delta \in (0,1/3)$,  
\begin{equation*}
\Phi_{n,r}(\mc{C}[X]) \leq \frac{1 + \delta}{1 - \delta} \cdot  \frac{\cut_{\Pbb,r}(\mc{C})}{\min\{\vol_{\Pbb,r}(\mc{C}), \vol_{\Pbb,r}(\mc{C}^c)\}} = \frac{1 + \delta}{1 - \delta} \cdot \Phi_{\Pbb,r}(\mc{C}) \leq (1 + 3\delta) \cdot \Phi_{\Pbb,r}(\mc{C})  
\end{equation*}
with probability at least $1 - 3\exp\{-n \delta^2 \cut_{\Pbb,r}(\mc{C})/4\}$. This establishes~\eqref{eqn:sample_to_population_normalized_cut} upon taking $b_1 := 3\cut_{\Pbb,r}(\mc{C})/4$.

\subsection{Sample-to-population: local spread}
\label{subsec:sample_to_population_local_spread}
In this subsection we establish~\eqref{eqn:sample_to_population_local_spread}. To ease the notational burden, let $\wt{G}_{n,r} := G_{n,r}\bigl[\mc{C}[X]\bigr]$. 
Conditional on~$\wt{n}$, it follows from Lemma~\ref{lem:bernstein_union} that with probability at least $1 - \wt{n}\exp\{-(\wt{n} - 1) \delta^2d_{\min}(\wt{\Pbb})/4\}$,
\begin{equation}
\label{pf:sample_to_population_local_spread_2}
(1 - \delta) \cdot d_{\min}(\wt{\Pbb}) \leq \frac{1}{\wt{n} - 1} d_{\min}(\wt{G}_{n,r}),
\end{equation}
Likewise it follows from Lemma~\ref{lem:hoeffding} that with probability at least $1 - \exp\{-\wt{n}\delta^2 \vol_{\wt{\Pbb},r}(\mc{C})/6\}$,
\begin{equation*}
(1 - \delta) \cdot \vol_{\wt{\Pbb},r}(\mc{C}) \leq \frac{1}{\wt{n}(\wt{n} - 1)} \vol(\wt{G}_{n,r}).
\end{equation*}
Finally, it follows from Lemma~\ref{lem:hoeffding_2} that with probability at least $1 - \exp\{-n\delta^2 \Pbb(\mc{C})/4\}$
\begin{equation}
\label{pf:sample_to_population_local_spread_1}
\wt{n} \geq (1 - \delta) \cdot n \cdot \Pbb(\mc{C}),
\end{equation}
and therefore by~\eqref{eqn:sample_to_population_local_spread_sample_complexity}, $(\wt{n} - 1)/\wt{n} \geq 1 - \delta$. Consequently for any $\delta \in (0,1/3)$,
\begin{equation*}
s_{n,r}(\mc{C}[X]) = \frac{d_{\min}(\wt{G}_{n,r})^2}{\vol(\wt{G}_{n,r})} = \frac{\wt{n} - 1}{\wt{n}} \cdot \frac{\frac{1}{(\wt{n} - 1)^2}d_{\min}(\wt{G}_{n,r})^2}{\frac{1}{\wt{n}(\wt{n} - 1)}\vol(\wt{G}_{n,r})} \geq \frac{(1 - \delta)^3}{(1 + \delta)} \cdot  \frac{d_{\min}(\wt{\Pbb})^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \geq (1 - 4\delta) \cdot s_{\Pbb,r}(\mc{C}).
\end{equation*}
with probability at least $1 - n\exp\{-n\Pbb(\mc{C}) \cdot \delta^2d_{\min}(\wt{\Pbb})/9\} - \exp\{-n\Pbb(\mc{C})\delta^2 \cdot \vol_{\wt{\Pbb},r}(\mc{C})/14\} - \exp\{-n\delta^2 \Pbb(\mc{C})/4\}$. This establishes~\eqref{eqn:sample_to_population_local_spread} upon taking $b_2 := \Pbb(\mc{C}) \cdot d_{\min}(\wt{\Pbb})/14$.

% AJG: To get the ``with high probability'' statement, I
%   (i) Used {eqn:sample_to_population_local_spread_sample_complexity} to deduce that \wt{n} - 1 \geq (1 - \delta)^2 \Pbb(\mc{C}) n;
%   (ii) replaced (1 - \delta) terms by factors of 2/3.

@@

In this section we establish~\eqref{eqn:sample_to_population_conductance}. As mentioned in our main text, the proof of~\eqref{eqn:sample_to_population_conductance} relies on a high-probability upper bound of the $\infty$-transportation distance between $\mbb{P}$ and $\mbb{P}_n$, from~\citep{garciatrillos16b}. We begin by reviewing this upper bound, which we restate in Theorem~\ref{thm:garciatrillos16}. Subsequently in Proposition~\ref{prop:conductance_lb_transportation_distance}, we relate the $\infty$-transportation distance between two measures $\mbb{Q}_1$ and $\mbb{Q}_2$ to the difference of their conductances. Together these results will imply~\eqref{eqn:sample_to_population_conductance}.

@@

\paragraph{$\infty$-transportation distance between empirical and population measures.}
We now review the relevant upper bound on $\Delta_{\infty}(\Pbb,\Pbb_n)$, which holds under the following mild regularity conditions.
\begin{enumerate}[label=(A\arabic*)]
	\item 
	\label{asmp:bounded_density} 
	The distribution $\Pbb$ has density $g: D \to (0,\infty)$ such that there exist $g_{\min} \leq 1 \leq g_{\max}$ for which
	\begin{equation*}
	(\forall x \in D)~~ g_{\min} \leq g(x) \leq g_{\max}.
	\end{equation*}
	\item 
	\label{asmp:domain} 
	The distribution $\Pbb$ is defined on a bounded, connected, open domain $D \subseteq \Rd$. If $d \geq 2$ then additionally $D$ has Lipschitz boundary.
\end{enumerate}
When $d = 1$, it follows from Proposition~6.2 of \textcolor{red}{Dudley} that $\Delta_{\infty}(\Pbb,\Pbb_n) \leq B_4 \|F - F_n\|_{\infty}$ for some positive constant $B_4$, and in turn from the DKW inequality (\textcolor{red}{Massart}) that
\begin{equation}
\label{eqn:ot_distance_1d}
\Delta_{\infty}(\Pbb,\Pbb_n) \leq B_4 \sqrt{\frac{\ln(2n/B_2)}{n}}
\end{equation}
with probability at least $1 - B_2/n$.

When $d \geq 2$, \cite{garciatrillos16b} derive an upper bound on the transportation distance $\Delta_{\infty}(\mbb{P},\mbb{P}_n)$.
\begin{theorem}[Theorem~1.1 of \cite{garciatrillos16}]
	\label{thm:garciatrillos16}
	Suppose $\Pbb$ satisfies~\ref{asmp:domain} and~\ref{asmp:bounded_density}. Then, there exists positive constants $B_{2}$ and $B_4$ that do not depend on $n$, such that with probability at least $1 - B_2/n$:
	\begin{equation*}
	\Delta_{\infty}(\mbb{P},\mbb{P}_n) \leq B_4 \cdot 
	\begin{dcases*}
	\frac{\ln(n)^{3/4}}{n^{1/2}},&~~\textrm{if $d = 2$,} \\
	\frac{\ln(n)^{1/d}}{n^{1/d}},&~~\textrm{if $d \geq 3$.}
	\end{dcases*}
	\end{equation*}
\end{theorem}
Assuming the candidate cluster $\mc{C}$ and conditional distribution $\wt{\Pbb}$ satisfy~\ref{asmp:domain} and~\ref{asmp:bounded_density}, then~\eqref{eqn:ot_distance_1d} ($d = 1)$ or Theorem~\ref{thm:garciatrillos16} ($d \geq 2$) apply to $\Delta_{\infty}(\wt{\Pbb},\wt{\Pbb}_n)$; we will use these upper bounds on $\Delta_{\infty}(\wt{\Pbb},\wt{\Pbb}_n)$ to show~\eqref{eqn:sample_to_population_conductance}.

@@

\subsection{Population functionals, hard case}
\label{subsec:density_cluster_hard_case}
Let $\Pbb$ be the hard case distribution over rectangular domain $\mc{X}$, defined as in~\eqref{eqn:lb_density}, and $\mc{L}$ the lower half of $\mc{X}$. Suppose $r \in (0,\sigma/2)$. Then the population normalized cut $\Phi_{\Pbb,r}(\mathcal{L})$ is upper bounded,
\begin{equation}
\label{eqn:ncut_lower_set}
\Phi_{\Pbb,r}(\mc{L}) \leq \frac{8}{3} \cdot \frac{r}{\rho}.
\end{equation}
and the population local spread $s_{\Pbb,r}(\mc{X})$ is lower bounded,
\begin{equation}
\label{eqn:local_spread_ub}
s_{\Pbb,r}(\mc{X}) \geq \frac{\pi r^2 \epsilon^2}{2 \rho\sigma}
\end{equation}
